Tabular Constraint Learning

### 2016-04-14 Paper/ECAI/tabular.tex
rewriting the paper

\tias{Use spreadsheet instead of Excel}
\tias{Spreadsheet is an instance of tabular data, but not the only one?}

@@

This makes tabular data understanding and constraint learning new and challenging, and it is not only challenging for a machine but for a human as well. Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the Spreadsheet risk\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaw in the famous economics papers \cite{flaw_excel} and caused billion-losses in the financial industry. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

\section{Formalization}
\samuel{Explain concepts on example}
\sergey{subgroup notation, rework the solution definition}
\subsection{Groups, Type-consistency and Constraints}
In this work we focus on the spreadsheet data, which is physically one big table, but conceptually consists of multiple tables, see Fig \ref{fig:main_example} for an example. Formally, a table is $n \times m$ matrix, each entry is called a \textit{cell}. A cell has a type, e.g. namely numeric. In this work we distinguish the two types of data: numeric and textual. The numeric type has two subtypes: integers and floats. We also consider the special element called \textit{None}, which has two types: numeric and textual.

@@

We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If the subgroup $g$ must constain only one vector, we write $g \in G$.

\tias{Use generic constraint}
A \textit{Spreadsheet constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. \textit{\CName} is the textual name of the constraint together with its variable names. \textit{\CSignature} is a set of constraints specifying the properties of the group assignments, corresponding to the variables in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. \textit{\CFunction} is a set of constraints specifying that the data in the subgroups satisfies the function. For example, a spreadsheet constraint \textit{rank} has \CName \textit{Y = RANK(X)}, where $X$ and $Y$ are the variables; its \CSignature is: the group $G_X$ (associated with X) is numeric, $G_Y$ is integer and the length of vectors in $G_X$ is the same as in $G_Y$; its \CFunction is the following constraint: a pair of vectors $X,Y$ is a solution iff $X \in G_X, Y \in G_Y$ and each value in $X$ has the rank (possibly with ties) specified in $Y$.

\samuel{2nd problem statement (from approach), remove EQUAL}

A \textit{tabular solution} $S$ to a spreadsheet constraint $C$ is a tuple $(X_1, \dots, X_n)$ where $n$ is the number of variables in $C$ and each $X_i$ is a subgroup in \groups, such that $(X_1,\dots,X_n)$ satisfies both the \CSignature and \CFunction of the constraint $C$.

@@

  In theory each constraint is independent and equally useful. In practice, however, things are different. Assume that for some $X$ both \textit{series(X)} and \textit{alldifferent(X)} hold. From a user's perspective the last solution is useless, since he or she already knows that series constraint implies alldifferent constraint. To capture this condensed representation \cite{condensed} of solutions, we have introduced a specificity (generality) order between constraints, which is depicted using solid (dashed) arrows in Figure \ref{fig:learning_order}. This matters from implementation perspective as well, if we discovered that $X$ is a not permutation, then $X$ should not be tested for the series constraint at all. A solid lines from $C_1$ to $C_2$ indicates that $C_2$ is \textit{more specific} than $C_1$ and if $X$ is not a solution to $C_1$, then it is not a solution to $C_2$. A dashed arrow from $C_1$ to $C_2$ indicates that $C_1$ is \textit{a more specific} constraint than $C_2$ and therefore if $X$ is a solution for $C_1$ it should not be tested for $C_2$.

@@

\paragraph{Example}
Consider the \ecfkey{FK}{PK} (foreign-key) constraint: for every solution the subgroup (vector) $G_{PK}$ assigned to $PK$ must fulfill \ecalldiff{G_{PK}}.
Therefore, we say indicate \textit{foreign-key} as being dependent on \textit{all-different} and the latter will be treated first.
The \ecrank{Y}{X} constraint does not have any dependencies, however in a restricted version without ties the vector $Y$ would always be a permutation of numbers $1, 2, ..., length(Y)$.

@@

\paragraph{Example}
The \ecrank{Y}{X} constraint requires $Y$ to be integer or textual and $X$ to be numeric.
This allows to filter candidate groups for these variables.
Additionally, $Y$ and~$X$ must be vectors of the same length, therefore, only groups with the same length are considered.
Groups that does not satisfy these requirements cannot contain subgroups (vectors) that satisfy \ecrank{Y}{X}.

@@

\paragraph{Example}
Consider a candidate assignment $G_X, G_Y$ associated with the variables $X,Y$ of \ecrank{Y}{X}, then \findassignment selects single vectors $G_X^{i}, G_Y^{j}$ such that \ecrank{G_Y^{j}}{G_X^{i}} is true.
For example, in Figure~\ref{fig:main_example} the group $G = \range{T_1}{\rangeall}{\rangeto{3}{8}}$ can serve as both $G_X$ and $G_Y$.
\findassignment could then select $G_X^{i}$ to be $\range{T_1}{\rangeall}{7}$ and $G_Y^{j}$ to be $\range{T_1}{\rangeall}{8}$ which forms a valid solution.
\\\\
For some constraints (most notably aggregates) there may be multiple overlapping subsets that satisfy the constraint.
We chose to omit subsets that are fully contained within a larger subset since this can lead to overfitting.
If the returned subsets are too large and accidentally contains unrelated vectors, the user can split the group to reflect this structure.

@@

Our system internally supports ASP (using Clingo 4.5.4 \cite{clingo}), Minizinc (Gecode-backend 2.0.2 \cite{minizinc}) and a native python CSP solver \cite{python_constraint}. All the constraints mentored in the paper can be run purely in the internal python engine without any external software. The future versions of the system will provide user interfaces to the ASP, Minizinc and Python-Constraint languages.

\subsection{Constraints}
% As emphasized before, the set of Excel constraints has a partial order in which they should be learned. For certain constraints this order does not matter, such as \textit{rank} or \textit{product}. For many others they have to be learned in the order, which is specified as a DAG in Figure \ref{fig:learning_order} (if a constraint is not in the graph, it is independent). This figure also specifies a potential parallelization of the computation, since each connected component is independent of the others.

@@

  \end{tabularx}
  \caption{Spreadsheet Constraints \sergey{Luc, we need your comments here on the notation of $X,Y$ and $G_X,G_Y$}}
  \label{table:constraints}
\end{table*}

@@

\subsection{Case study}
Let us illustrate our implementation using the example presented in figure~\ref{fig:main_example}.
This example combines several smaller examples that were used in an exercise session to teach Excel into one spreadsheet.
Figure~\ref{fig:sol_example} shows the constraints that we expect to find.

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
%
      & \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      & \ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecaggc{AVERAGE}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MAX}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MIN}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggif{SUM}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \ecaggif{MAX}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints case study}
\end{figure}

@@

\subsubsection*{Q1. How accurate is the approach?}
Fortunately, our system is currently able to find $100\%$ of the expected constraints (90) on the benchmark suite.
There are only three essential constraints that have not been implemented yet, therefore our system currently identifies $96.77\%$ of the constraints that we hope to discover.

@@

\subsection{Insights}
We should explain Figure \ref{fig:fbi}, e.g. if we aggregate by the crime now, the missing value would not get into statistics, which seems faulty, since we can directly compute it.

>>>>

@@

This makes tabular data understanding and constraint learning new and challenging, and it is not only challenging for a machine but for a human as well. Due to the complexity of numeric computations in spreadsheets, people often fail to grasp the properties of the data, which leads to so called the \textit{Spreadsheet risk}\footnote{\url{https://en.wikipedia.org/wiki/Spreadsheet\#Spreadsheet_risk}} and even caused major flaws in the famous economics papers \cite{flaw_excel} and billion-losses in the financial industry \cite{spreadsheet_risk_loss}. Tabular constraint learning provides a potential cure by indicating learned constraint violations and suggest possible fixes.

@@

\section{Formalization}
\samuel{Explain concepts on example}
\sergey{subgroup notation, rework the solution definition}
\subsection{Groups, Type-consistency and Constraints}
In this work we focus on the spreadsheet data, which is physically one big table, but conceptually consists of multiple tables, see Figure \ref{fig:main_example} for an example. Formally, a table is $n \times m$ matrix, each entry is called a \textit{cell}. A cell has a type, e.g. namely numeric. In this work we distinguish the two types of data: numeric and textual. The numeric type has two subtypes: integers and floats. We also consider the special element called \textit{None}, which has two types: numeric and textual.

@@

We refer to a subgroup of a group $G$ in \groups as $g$. A \textit{subgroup} $g$ is a subrange of vectors in a group $G$. If a subgroup $g$ must constain only one vector, we write $g \in G$.

A \textit{Spreadsheet constraint} is a triple \textit{(\CName, \CSignature, \CFunction)}. Let us elaborate on each of them. \textit{\CName} is the textual name of the constraint together with the list of its variables $v_1,\dots,v_n$. \textit{\CSignature} is a constraint specifying the properties of the group assignments, corresponding to the variables $v_1,\dots,v_n$ in \CName, such as their types, e.g., requiring them to be integers or constraining the sizes, e.g. the length of vectors in the arguments must be equal. These are constraints on the group meta-information not on the actual group content. \textit{\CFunction} is a constraint specifying that the data in the subgroups satisfies the function. For example, a spreadsheet constraint \textit{rank} has \CName \textit{$v_y$ = RANK($v_x$)}, where $v_x$ and $v_y$ are the variables; its \CSignature is: the subgroup assigned to $v_x$ must be numeric, $v_y$ must be integer and the length of vectors in $v_x$ is the same as in $v_y$; its \CFunction is the following constraint: the mapping $\{ v_x \mapsto g_x, v_y \mapsto g_y \}$ is a solution iff $g_x \in G_x, g_y \in G_y$ ($G_x \in \groups, G_y \in \groups$) and each value in $g_x$ has the rank (possibly with ties) specified in $g_y$.

Let us now define what is a tabular solution for a spreadsheet constraint in general. Let $C$ denote a spreadsheet constraint, $S$ a mapping from variables to the subgroups of \groups, and $C^S$ the constraint $C$, in which all variables are substituted with the corresponding subgroups from $S$, then $S$ is a \textit{tabular solution} to $C$ iff both $\CSignature^S$ and $\CFunction^S$ hold.

@@

  In theory each constraint is independent and equally useful. In practice, however, things are different. Assume that for some $g_x$ both \textit{series($g_x$)} and \textit{alldifferent($g_x$)} hold. From a user's perspective the last solution is useless, since he or she already knows that the series constraint implies the alldifferent constraint. To capture this condensed representation \cite{condensed} of solutions, we have introduced a specificity (generality) order between constraints, which is depicted using solid (dashed) arrows in Figure \ref{fig:learning_order}. This matters from implementation perspective as well, if we discovered that $g_x$ is not a permutation, then $g_x$ should not be tested for the series constraint at all. A solid lines from $C_1$ to $C_2$ indicates that $C_2$ is \textit{more specific} than $C_1$ and if $S$ is not a solution to $C_1$, then it is not a solution to $C_2$. A dashed arrow from $C_1$ to $C_2$ indicates that $C_1$ is \textit{a more specific} constraint than $C_2$ and therefore if $S$ is a solution for $C_1$ it should not be tested for $C_2$.

@@

\paragraph{Example}
Consider the \ecfkey{FK}{PK} (foreign-key) constraint: for every solution the subgroup (vector) $v_{PK}$ assigned to \textit{PK} must fulfill \ecalldiff{v_{PK}}.
Therefore, we say indicate \textit{foreign-key} as being dependent on \textit{all-different} and the latter will be treated first.
The \ecrank{v_y}{v_x} constraint does not have any dependencies, however in a restricted version without ties the vector $v_y$ would always be a permutation of numbers $1, 2, ..., \textit{length}(v_y)$.

@@

\paragraph{Example}
The \ecrank{v_y}{v_x} constraint requires $v_y$ to be integer or textual and $v_x$ to be numeric.
This allows to filter candidate groups for these variables.
Additionally, $v_y$ and~$v_x$ must be vectors of the same length, therefore, only groups with the same length are considered.
Groups that does not satisfy these requirements cannot contain subgroups (vectors) that satisfy \ecrank{v_y}{v_x}.

@@

\paragraph{Example}
Consider a candidate assignment $G_x, G_y$ associated with the variables $v_x,v_y$ of \ecrank{v_y}{v_x}, then \findassignment selects single vectors $g_y \in G_y, g_x \in G_x$ such that \ecrank{g_y}{g_x} is true, i.e. it searches for a mapping $S$ where $v_x$ ($v_y$) must be mapped to a subgroup of $G_x$ ($G_y$).
For example, in Figure~\ref{fig:main_example} the group $G = \range{T_1}{\rangeall}{\rangeto{3}{8}}$ can serve as both $G_x$ and $G_y$.
Then \findassignment could select $g_x$ to be $\range{T_1}{\rangeall}{7}$ and $g_y$ to be $\range{T_1}{\rangeall}{8}$ which forms a valid solution.
\\\\
For some constraints (most notably aggregates) there may be multiple overlapping subsets that satisfy the constraint.
We chose to omit subsets that are fully contained within a larger subset since this can lead to overfitting.
If the returned subsets are too large and accidentally contains unrelated vectors, the user can split the group to reflect this structure.

@@

Our system internally supports ASP (using Clingo 4.5.4 \cite{clingo}), Minizinc (Gecode-backend 2.0.2 \cite{minizinc}) and a native Python CSP solver \cite{python_constraint}. All the constraints mentored in the paper can be run purely in the internal Python engine without any external software. The future versions of the system will provide user interfaces to the ASP, Minizinc and Python-Constraint languages to allow users to add new constraints in a declarative way.

\subsection{Constraints}

@@

  \end{tabularx}
  \caption{Spreadsheet Constraints \sergey{Here a detailed explanation on what is going on, what is essential, what is not? etc should be self-explanatory}}
  \label{table:constraints}
\end{table*}

@@

\subsection{Case study}
Let us illustrate our implementation using the example presented in Figure~\ref{fig:main_example}.
This example combines several smaller examples that were used in an exercise session to teach Excel into one spreadsheet.
Figure~\ref{fig:sol_example} shows the constraints that we expect to find.

\begin{figure}
  \label{fig:sol_example}
  {\small
    \begin{align*}
      & SERIES(\range{T_{1}}{\rangeall}{1}) \\
%
      & \ecrank{\range{T_{1}}{\rangeall}{8}}{\range{T_{1}}{\rangeall}{7}} \\
%
      & \ecsumc{\range{T_{2}}{1}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecsumc{\range{T_{6}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecsumr{\range{T_{1}}{\rangeall}{7}}{\range{T_{1}}{\rangeall}{\rangeto{3}{6}}} \\
%
      & \ecaggc{AVERAGE}{\range{T_{2}}{2}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MAX}{\range{T_{2}}{3}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggc{MIN}{\range{T_{2}}{4}{\rangeall}}{\range{T_{1}}{\rangeall}{\rangeto{3}{7}}} \\
%
      & \ecaggif{SUM}{\range{T_{1}}{\rangeall}{10}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \ecaggif{MAX}{\range{T_{1}}{\rangeall}{11}}{\range{T_{5}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}}{\range{T_{5}}{\rangeall}{2}} \\
%
      & \eclookup{\range{T_{4}}{\rangeall}{3}}{\range{T_{4}}{\rangeall}{2}}{\range{T_{1}}{\rangeall}{1}}{\range{T_{1}}{\rangeall}{2}} \\
%
      & \range{T_{6}}{\rangeall}{4} = PREV(\range{T_{6}}{\rangeall}{4}) + \range{T_{6}}{\rangeall}{2} - \range{T_{6}}{\rangeall}{3}
    \end{align*}
  }
  \caption{Expected constraints in the case study \sergey{in Figure \ref{fig:main_example}; some more explanation needed, not self explanatory}}
\end{figure}

@@

\subsubsection*{Q1. How accurate is the approach?}
\sergey{that's confusing, it can do 100\% but then all of a sudden 96\%, short explanation needed}
Fortunately, our system is currently able to find $100\%$ of the expected constraints (90) on the benchmark suite.
There are only three essential constraints that have not been implemented yet, therefore our system currently identifies $96.77\%$ of the constraints that we hope to discover.

@@

\subsection{Insights}
\sergey{We should explain Figure \ref{fig:fbi}, e.g. if we aggregate by the crime now, the missing value would not get into statistics, which seems faulty, since we can directly compute it.}
